Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-02 doc/papers/cgo2014/paper.tex
Comments + typos

As introduce in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
to evaluate these optimizations. It is use with Pythran (commit: 488b7ba)
with optimization enabled/disabled. Finally, we compare result with Pypy (???),
Parakeet (commit: 4e55810), Numba (commit: b8a79c5) and Python 2.7.

Benchmarking is done one two architecture with different backend compilers.

The first one holds 8 AMD Magny Cours processors for a
total of 48 cores. Each core has access to 64\,KB of L1 cache,
512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 6 cores of a processor. This configuration
provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
up to SSE4. We will refer to this configuration as \textbf{AMD48}.
On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to benefits
from OpenMP improvement. This clang version is based on clang 3.4 release
(LLVM:233b1e3f034, clang-omp:6f358bcd87).

The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
8 cores (multithreading is not used). Each node can access to  64\,KB of L1 cache,
256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 8 processors. This configuration
provides a total of 63\,GB of main memory. It supports up to AVX. We will
refer to this configuration as \textbf{Intel8}.
On Intel8, backend compilers is gcc 4.9 (20140528).

@@

% For performance comparison between different Pythran version
To compare the different optimization schemes, we run the numpy-benchmarks test 
suite multiple time on the \textit{rosen} benchmark.
Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
The difference between the two architectures can be explain by AVX support on 
the Intel8 architecture while AMD48 support up to SSE4 only.
% Haaaaa, I don't understand why OMP is not use...
% tODO : I have to redo mesures without forward substitution to show difference.

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{rosen_amd48.png}
\caption{Rosen on AMD48}
\label{fig:rosenamd48}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{rosen_intel8.png}
\caption{Rosen on Intel8}
\label{fig:rosenintel8}
\end{figure}

@@

Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
use of technique described in this paper. As we can see, both SIMD instruction
and parallelization have an impact on execution time. We can notice that SIMD
contribution is not as good as it could be on AMD48 as it doesn't have AVX support.
Also, we can notice that gcc as a backend compiler have more optimization opportunity.

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{arc_distance_amd48.png}
\caption{Arc distance on AMD48}
\label{fig:arc_distance}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{arc_distance_intel8.png}
\caption{Arc distance on Intel8}
\label{fig:arc_distance_intel8}
\end{figure}

At the opposite, Figure~\ref{fig:diffusion} show that clang is better when generated
code make a more intensive use of templates (heavily use with gsliced\_ndarray).
gsliced\_ndarray is a non aligned view on an array in this case so SIMD optimization
can't be applied on the second dimension even if loop fusion and temporary
elimination is performed.  We can notice that Parakeet perform really well on
extensive gslice use.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{diffusion_amd48.png}
\caption{Diffusion on AMD48}
\label{fig:diffusion}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{diffusion_intel8.png}
\caption{Diffusion on Intel8}
\label{fig:diffusion_intel8}
\end{figure}

Finally, Figure~\ref{fig:l2norme} show the l2norme algorithm introduce in
\S~\ref{sec:stackoverflow}. Comparison with other compiler show that Pythran approach
is correct for these kind of functions.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{l2norm_amd48.png}
\caption{L2norm on AMD48}
\label{fig:l2norme}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{l2norm_intel8.png}
\caption{L2norm on Intel8}
\label{fig:l2norm_intel8}
\end{figure}

>>>>

As introduced in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
to evaluate the optimizations described in\S~\ref{sec:pythran}. It uses Pythran (commit: 488b7ba)
with optimization enabled/disabled. The results are compared with Pypy \todo{add}(???),
Parakeet (commit: 4e55810), Numba (commit: b8a79c5) and Python 2.7.

Benchmarking is done on two architectures with different backend compilers. The
\texttt{timeit} module is used to perform the measurments, taking the average
execution time over several runs.

The first one holds 8 AMD Magny Cours processors for a
total of 48 cores. Each core has access to 64\,KB of L1 cache,
512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 6 cores of a processor. This configuration
provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
up to SSE4. We refer to this configuration as \textbf{AMD48}.
On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to benefits
from OpenMP improvement. This clang version is based on clang 3.4 release
(LLVM:233b1e3f034, clang-omp:6f358bcd87).

%FIXME: je me demande si on devrait pas utiliser juste cette archi. Qu'est ce que l'autre apporte ? On sous utilise complètement les 48 coeurs...
The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
8 cores (multithreading is not used). Each node can access to  64\,KB of L1 cache,
256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 8 processors. This configuration
provides a total of 63\,GB of main memory. It supports up to AVX. We
refer to this configuration as \textbf{Intel8}.
On Intel8, backend compilers is gcc 4.9 (20140528).

@@

% For performance comparison between different Pythran version
To compare the different optimization schemes, we run the numpy-benchmarks test
suite multiple time on the \textit{rosen} benchmark.
Figures~\ref{fig:rosenamd48} and~\ref{fig:rosenintel8} illustrate the results.
The difference between the two architectures can be explain by AVX support on
the Intel8 architecture while AMD48 supports up to SSE4 only.
% Haaaaa, I don't understand why OMP is not use...
% tODO : I have to redo mesures without forward substitution to show difference.
% SG: may be the number of iteration is not enough? what array size did you use? (it should be specified in the text secion, BTW

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{rosen_amd48.png}
\caption{Execution time of the rosen kernel on AMD48.}
\label{fig:rosenamd48}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=.5\textwidth]{rosen_intel8.png}
\caption{Execution time of the rosen kernel on Intel8.}
\label{fig:rosenintel8}
\end{figure}

@@

Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
use of the techniques described in this paper. As we can see, both SIMD instruction
and parallelization have an impact on execution time. We can notice that SIMD
contribution is not as good as it could be on AMD48 as it doesn't have AVX support.
Also, we can notice that gcc as a backend compiler finds more optimization opportunities.
%FIXME: mehdi propose de n'utiliser qu'un backend pour centrer le propos sur Pythran/Numpy.
% ça allègera aussi les graphes...

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{arc_distance_amd48.png}
\caption{Execution time of the arc distance kernel on AMD48.}
\label{fig:arc_distance}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{arc_distance_intel8.png}
\caption{Execution time of the arc distance kernel on Intel8.}
\label{fig:arc_distance_intel8}
\end{figure}

At the opposite, Figure~\ref{fig:diffusion} shows that clang is better when generated
code make a more intensive use of templates (heavily used with \texttt{gsliced\_ndarray}).
\texttt{gsliced\_ndarray} is a non aligned view on an array in this case so SIMD optimization
can't be applied on the second dimension even if loop fusion and temporary
elimination is performed.  We can notice that Parakeet performs really well on
extensive gslice use.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
%SG: +1

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{diffusion_amd48.png}
\caption{Execution time of the diffusion kernel on AMD48.}
\label{fig:diffusion}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{diffusion_intel8.png}
\caption{Execution time of the diffusion kernel on Intel8.}
\label{fig:diffusion_intel8}
\end{figure}

Finally, Figure~\ref{fig:l2norme} show the l2norme algorithm introduce in
\S~\ref{sec:stackoverflow}. Comparison with other compiler show that Pythran approach
is correct for these kind of functions.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
%SG: +1
\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{l2norm_amd48.png}
\caption{Execution time of the l2norm kernel on AMD48.}
\label{fig:l2norme}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{l2norm_intel8.png}
\caption{Execution time of the l2norm kernel on Intel8.}
\label{fig:l2norm_intel8}
\end{figure}
